---
slug: "join"
title: "Join two data sets"
date: 2021-12-24T13:26:57-05:00 
publishdate: 2021-12-24
lastmod: 2021-12-24
tags: ["rtlas"]
draft: false
autonumbering: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
# Options to have images saved in the post folder
# And to disable comment symbols before each line of output
knitr::opts_chunk$set(
echo = FALSE,
message = FALSE,
results = "hold",
fig.show = "hold",
fig.path = "",
comment = "")

# Knitr hook to use Hugo markdown render image
knitr::knit_hooks$set(plot = function(x,options) {
    base = knitr::opts_knit$get('base.url')
    if (is.null(base)) base = ''
    alt = ifelse (is.null(options$alt),"",options$alt)
    cap = ifelse (is.null(options$caption),"",options$caption)
    if (alt != ""){
      sprintf('![%s](%s%s "%s")', alt, base, x, cap)
    } else {
      sprintf('![%s](%s%s)', cap, base, x)  
      }
})

# knitr hook to use Hugo highlighting options
knitr::knit_hooks$set(
source = function(x, options) {
hlopts <- options$hlopts
  paste0(
    "```r ",
    if (!is.null(hlopts)) {
    paste0("{",
      glue::glue_collapse(
        glue::glue('{names(hlopts)}={hlopts}'),
        sep = ","
      ), "}"
      )
    },
    "\n", glue::glue_collapse(x, sep = "\n"), "\n```\n"
  )
}
)
```

```{r include=FALSE}
source("../../../static/r/script.R")
```

```{r chunkname, class="figure", eval=FALSE, results="hide", fig.retina=2, out.width="100", alt="Label. alternative text please make it informative", caption="this is what this image shows, write it here or in the paragraph after the image as you prefer"}
plot(1:10)
```

## Problem

`marks.xlsx` has the following columns:

```{r include=FALSE}
# simulate a dataset with ID, names, mid-term marks, and final marks
marks <- tibble::tibble(
  id = sample(1:100, size = 30, replace = FALSE),
  midterm = sample(5:50, size = 30, replace = TRUE),
  final = sample(5:50, size = 30, replace = TRUE),
  lab1 = rep(NA_integer_, times = 30),
  lab2 = rep(NA_integer_, times = 30),
  lab3 = rep(NA_integer_, times = 30)
) |>
  dplyr::arrange(id)
# simulate a dataset with ID, names, lab1-3 marks
labs <- tibble::tibble(
  id = c(1:100),
  lab1 = sample(0:10, size = 100, replace = TRUE),
  lab2 = sample(0:10, size = 100, replace = TRUE),
  lab3 = sample(0:10, size = 100, replace = TRUE)
)
```

```{r}
marks
```


Columns `lab1`, `lab2`, `lab3` are currently empty.
Their values are stored in another file `labs.xlsx`,

```{r}
labs
```

## Solution

The students included in `marks.xlsx` is a subset of `labs.xlsx`,
evidenced by their different number of rows.
To grab the `lab1` - `lab3` data from `labs.xlsx` and fill them in `marks.xlsx`,
I can use `dplyr::left_join()`.

```{r echo=TRUE}
marks <- dplyr::left_join(marks[, 1:3], labs, by = "id")
marks
```

I can then export `marks`

```{r eval=FALSE, echo=TRUE}
readr::write_csv(tmp, file = "path/to/marks_new.csv")
```

## Epilogue

When I implemented this solution,
there was one more complication:
the `labs.xlsx` was originally not a single file,
but separated into 9 files.
I imported and concatenated them in one go with `purrr`.


```{r eval=FALSE, echo=TRUE}
# assuming the the 9 files are names "lab#.xlsx"
labs_file <- list.files(path = "./marks", pattern = "^lab", full.names = TRUE)
labs_data <- purrr::map_dfr(labs_file, ~readxl::read_excel(.x))
```

