---
slug: "svd4"
title: "Singular Value Decomposition - Eigenvectors of A Symmetric Matrix"
date: 2021-11-07T17:41:05-05:00 
publishdate: 2021-11-08
lastmod: 2021-11-07
tags: ["matrix"]
draft: false
autonumbering: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
# Options to have images saved in the post folder
# And to disable comment symbols before each line of output
knitr::opts_chunk$set(echo=FALSE, message=F, fig.path = "", comment = "")

# Knitr hook to use Hugo markdown render image
knitr::knit_hooks$set(plot = function(x,options) {
      base = knitr::opts_knit$get('base.url')
      if (is.null(base)) base = ''
      alt = ifelse (is.null(options$alt),"",options$alt)
      cap = ifelse (is.null(options$caption),"",options$caption)
      if (alt != ""){
        sprintf('![%s](%s%s "%s")', alt, base, x, cap)
      } else {
        sprintf('![%s](%s%s)', cap, base, x)  
        }
  })

# knitr hook to use Hugo highlighting options
knitr::knit_hooks$set(
  source = function(x, options) {
  hlopts <- options$hlopts
    paste0(
      "```r ",
      if (!is.null(hlopts)) {
      paste0("{",
        glue::glue_collapse(
          glue::glue('{names(hlopts)}={hlopts}'),
          sep = ","
        ), "}"
        )
      },
      "\n", glue::glue_collapse(x, sep = "\n"), "\n```\n"
    )
  }
)
```

```{r include=FALSE}
source("../../../static/r/script.R")
```

```{r include=FALSE}
eigen_transform <- function(matrix,
  mat = "a", l1, l2, l3, l4, r1, r2, r3, r4) {
  # calculate eigenvalues and eigenvectors for `matrix`
  eigen <- eigen(matrix)

  u_1 <- eigen$vectors[, 1]
  u_2 <- eigen$vectors[, 2]

  # prepare the drawing canvas
  # evenly split the canvas into left and right
  oldpar <- par(mfrow = c(1, 2))

  empty_canvas(main = "Original vectors")
  unit_circle() |> plot_unit_circle()
  # draw two vectors u1 and u2
  matlib::vectors(rbind(u_1, u_2),
    col = c("red", "red"),
    lwd = c(2, 2),
    angle = 15,
    labels = NULL
  )
  text(l1, l2, expression(bold(u[1])))
  text(l3, l4, expression(bold(u[2])))

  # transformation matrix
  mat_trans <- matrix

  # multiply mat_trans with each vector from vec_stack,
  # effectively transforming the circule represented by vec_stack
  # Not a transparent solution, to optimize
  empty_canvas(main = "New vectors after transformation")
  transform(mat_trans) |> plot_transformed(mat = mat, xpos = 1, ypos = 2.7)

  t_1 <- mat_trans %*% u_1
  t_2 <- mat_trans %*% u_2

  matlib::vectors(rbind(t(t_1), t(t_2)),
    col = c("red", "red"),
    lwd = c(2, 2),
    angle = 15,
    labels = NULL
  )
  text(r1, r2, bquote(paste(bold(.(mat)), bold(u[1]), " = ", lambda, bold(u[1]))))
  text(r3, r4, bquote(paste(bold(.(mat)), bold(u[2]), " = ", lambda, bold(u[2]))))

  # restore default parameter of canvas
  par(oldpar)
  }
```



In this fourth post of the [series]({{< relref "../svd/index.md" >}}),
let's first review the two concepts
introduced in the [last post]({{< relref "../svd3/index.md" >}}) ---
transpose and dot product --- but in a new context.

## Partitioned matrix

When calculating the transpose of a matrix, 
it is usually useful to represent it as a partitioned matrix.
For example, the matrix

$$
\mathbf{C} = 
\begin{bmatrix}
  5 & 4 & 2 \\
  7 & 1 & 9 \\
\end{bmatrix}
$$ 

can be also written as:

$$
\mathbf{C} =
\begin{bmatrix}
 {\mathbf{u}}_1  & {\mathbf{u}}_2 & {\mathbf{u}}_3 \\
\end{bmatrix}
$$ 

where 

$$
\begin{equation*}
  \mathbf{u}_1 = 
  \begin{bmatrix}
    5 \\
    7 \\
  \end{bmatrix},~
  \mathbf{u}_2 = 
  \begin{bmatrix}
    4 \\
    1 \\
  \end{bmatrix},~
  \mathbf{u}_3 = 
  \begin{bmatrix}
    2 \\
    9 \\
  \end{bmatrix}
\end{equation*}
$$ 

We can think of each column of $\mathbf{C}$ as a column vector, 
and $\mathbf{C}$ can be thought of as a matrix with just one row.
To write the transpose of $\mathbf{C}$, 
we can simply turn this row into a column, 
similar to what we do for a row vector. 
The only difference is that each element in $\mathbf{C}$ is now a vector itself
and should be transposed too.

$$
{\mathbf{u}_1}^\top =
\begin{bmatrix}
  5 & 7 \\
\end{bmatrix},~
{\mathbf{u}_2}^\top =
\begin{bmatrix}
  4 & 1 \\
\end{bmatrix},~
{\mathbf{u}_3}^\top =
\begin{bmatrix}
  2 & 9 \\
\end{bmatrix},~
$$ 

Therefore, 

$$
\mathbf{C}^\top = 
\begin{bmatrix}
  {\mathbf{u}_1}^\top \\
  {\mathbf{u}_2}^\top \\
  {\mathbf{u}_3}^\top \\
\end{bmatrix} = 
\begin{bmatrix}
  5 & 7 \\
  4 & 1 \\
  2 & 9 \\
\end{bmatrix}
$$ 

Each row of $\mathbf{C}^\top$ is the transpose of the corresponding column 
of the original matrix $\mathbf{C}$.

Let matrix $\mathbf{A}$ be a partitioned column matrix 
and matrix $\mathbf{B}$ be a partitioned row matrix: 

$$
\mathbf{A} = 
\begin{bmatrix}
  \mathbf{a}_1 & \mathbf{a}_2 & \cdots & \mathbf{a}_p \\
\end{bmatrix},~
\mathbf{B} = 
\begin{bmatrix}
  {\mathbf{b}_1}^\top \\
  {\mathbf{b}_2}^\top \\
  \vdots\\
  {\mathbf{b}_p}^\top \\
\end{bmatrix}
$$ 

where each column vector $\mathbf{a}_i$ is defined as the $i$-th column of $\mathbf{A}$: 

$$
\begin{equation*}
  \mathbf{a}_i =
    \begin{bmatrix}
      a_{1,i} \\
      a_{2,i} \\
      \vdots \\\\
      a_{m,i} \\
    \end{bmatrix}
\end{equation*}
$$

For each element, the first subscript refers to the row number 
and the second subscript to the column number. 
Therefore, $\mathbf{A}$ is an $m \times p$ matrix. 
In addition, $\mathbf{B}$ is a $p \times n$ matrix 
where each row vector in ${\mathbf{b}_i}^\top$ is the $i$-th row of 
$\mathbf{B}$:

$$
\begin{equation*}
  \mathbf{b}_i^\top=
    \begin{bmatrix}
      b_{i1} & b_{i2} & \cdots & b_{in} \\
    \end{bmatrix}
\end{equation*}
$$ 

Note that by convention, a vector is written as a column vector. 
To write a row vector, 
we write it as the transpose of a column vector. 
$\mathbf{b}_i$ is a column vector, 
and its transpose is a row vector that captures the $i$-th row of $\mathbf{B}$. 
To calculate $\mathbf{A}\mathbf{B}$ : 

$$
\begin{equation*}
  \mathbf{C} = \mathbf{A}\mathbf{B} = 
  \begin{bmatrix}
    \mathbf{a}_1 & \mathbf{a}_2 & \cdots & \mathbf{a}_p \\
  \end{bmatrix}
  \begin{bmatrix}
    {\mathbf{b}_1}^\top \\
    {\mathbf{b}_2}^\top \\
    \vdots\\
    {\mathbf{b}_p}^\top \\
  \end{bmatrix} = 
  \cdots = 
  \mathbf{a}_1{\mathbf{b}_1}^\top + \mathbf{a}_2{\mathbf{b}_2}^\top + \cdots + \mathbf{a}_p{\mathbf{b}_p}^\top
\end{equation*}
$$

The product of the $i$-th column of $\mathbf{A}$ and the $i$-th row of $\mathbf{B}$ 
gives an $m \times n$ matrix, 
and all these matrices are added together to give $\mathbf{A}\mathbf{B}$ 
which is also an $m \times n$ matrix. 
As a special case, suppose that $\mathbf{x}$ is a column vector.
We can calculate $\mathbf{A}\mathbf{x}$ such that: 

$$
\mathbf{A}\mathbf{x} = 
\begin{bmatrix}
  \mathbf{a}_1 & \mathbf{a}_2 & \cdots & \mathbf{a}_p \\
\end{bmatrix}
\begin{bmatrix}
  x_1 \\
  x_2 \\
  \vdots \\
  x_p \\
\end{bmatrix} = 
x_1\mathbf{a}_1 + x_2\mathbf{a}_2 + \cdots + x_p\mathbf{a}_p
$$ 

$\mathbf{A}\mathbf{x}$ is simply a linear combination of the columns of $\mathbf{A}$. 

To calculate the dot product of two vectors $\mathbf{a}$ and $\mathbf{b}$ in `R`, 
we can use `a %*% b`, as matrix multiplication.

## Length of a Vector

Now that we are familiar with the transpose and dot product, 
we can define the length (also called the 2-norm) of vector $\mathbf{u}$ as: 

$$
\|\mathbf{u}\| = 
\sqrt{\mathbf{u}\cdot\mathbf{u}} = 
\sqrt{\mathbf{u}^\top\mathbf{u}} = 
\sqrt{u_1^2 + u_2^2 + \cdots + u_n^2}
$$ 

To normalize a vector $\mathbf{u}$, we simply divide it by its length 
to have the normalized vector $\mathbf{n}$ : 

$$
\mathbf{n} = \frac{\mathbf{u}}{\|\mathbf{u}\|}
$$ 

The normalized vector $\mathbf{n}$ is still in the same direction of $\mathbf{u}$, 
but its length is 1. 
Now we can normalize the eigenvector of $\lambda = -1$ that we saw before
in the [second post]({{< relref "../svd2/index.md" >}}) of this series:

$$
\begin{equation*}
  \mathbf{u}_2 = 
    \begin{bmatrix}
      -1 \\
      1 \\
    \end{bmatrix}
\end{equation*},
$$
$$
\begin{equation*}
  \|\mathbf{u}_2\| = \sqrt{(-1)^2 + 1^2} = \sqrt{2}
\end{equation*},
$$
$$
\begin{equation*}
  \mathbf{n} = \frac{\mathbf{u}_2}{\|\mathbf{u}_2\|} = 
  \begin{bmatrix}
    \frac{-1}{\sqrt{2}} \\
    \frac{1}{\sqrt{2}} \\
  \end{bmatrix} \approx 
  \begin{bmatrix}
    -0.7071 \\
    0.7071 \\
  \end{bmatrix}
\end{equation*}
$$ 

which is the same as the output of `eigen(mat_b)`.
As shown before, if we multiply (or divide) an eigenvector by a constant, 
the new vector is still an eigenvector for the same eigenvalue.
Therefore, by normalizing an eigenvector corresponding to an eigenvalue, 
we'd still have an eigenvector for that eigenvalue.

## Revisit Eigenvectors

Why are eigenvectors important to us? 
As mentioned before, an eigenvector simplifies the matrix multiplication 
into a scalar multiplication. 
In addition, they have some more interesting properties. 
Let me go back to matrix $\mathbf{A}$ 
which was used in a [previous post]({{< relref "../svd/index.md" >}})
and calculate its eigenvectors: 

$$
\mathbf{A} = 
\begin{bmatrix}
  3 & 2 \\
  0 & 2 \\
\end{bmatrix}
$$ 

In the [previous post]({{< relref "../svd/index.md" >}}),
this matrix transformed a set of vectors $\mathbb{X}$ forming a circle 
into a new set $\mathbb{T}$ forming an ellipse. 
Let's use `eigen()` to calculate its eigenvectors.

```{r echo=TRUE}
mat_a <- matrix(c(3, 0, 2, 2), 2)
eigen_a <- eigen(mat_a)
eigen_a
```

We got two eigenvectors: 

$$
\mathbf{u}_1 = 
\begin{bmatrix}
  1 \\
  0 \\
\end{bmatrix},~
\mathbf{u}_2 = 
\begin{bmatrix}
  -0.8944 \\
  0.4472 \\
\end{bmatrix}
$$ 

and the corresponding eigenvalues are: 

$$
\lambda_1 = 3,~\lambda_2 = 2
$$ 

Now we plot $\mathbb{X}$ and the eigenvectors of $\mathbf{A}$ --- 
$\mathbf{u}_1$ and $\mathbf{u}_2$, 
as well as vectors in $\mathbb{X}$ transformed by $\mathbf{A}$.

```{r eigenvectors, class="figure", caption="Eigenvectors (left) and transformed eigenvectors (right).", alt="eigenvectorsa. Eigenvectors (left) and transformed eigenvectors (right).", fig.retina=2}
eigen_transform(matrix = mat_a,
  mat = "A",
  1.5, 0,
  -0.4, 0.5,
  2, -0.4,
  -1.6, 0.1
)
```

Every vector in $\mathbb{X}$ (left), once transformed by $\mathbf{A}$, 
is stretched **and** rotated in $\mathbf{A}\mathbf{x}$ (Figure [1](#eigenvectors-figure) right),
except two vectors --- $\mathbf{u}_1$ and $\mathbf{u}_2$; 
those are only stretched, as though they are multiplied by a scalar. 
This is because they are the eigenvectors of $\mathbf{A}$, 
and multiplying a matrix with its eigenvector 
is equivalent to multiplying the eigenvector with its corresponding eigenvalue.

$$
\mathbf{A}\mathbf{u} = \lambda\mathbf{u}
$$ 

Let's try another matrix:

$$
\mathbf{B} = 
\begin{bmatrix}
  3 & 1 \\
  1 & 2 \\
\end{bmatrix}
$$ 

It's two eigenvectors: 

$$
\mathbf{u}_1 = 
\begin{bmatrix}
  0.8507 \\
  0.5257 \\
\end{bmatrix},~
\mathbf{u}_2 = 
\begin{bmatrix}
  -0.5257 \\
  0.8507 \\
\end{bmatrix}
$$ 

and the corresponding eigenvalues are:

$$
\lambda_1 = 3.618,~\lambda_2 = 1.382
$$ 

Figure [2](#eigenvector-figure) shows $\mathbb{X}$ and the eigenvectors of $\mathbf{B}$ --- 
$\mathbf{u}_1$ and $\mathbf{u}_2$, 
as well as vectors in $\mathbb{X}$ transformed by $\mathbf{B}$.

```{r eigenvectors-matrix-b, class="figure", caption="Eigenvectors (left) and transformed eigenvectors (right) for matrix B.", alt="eigenvectorsb. Eigenvectors (left) and transformed eigenvectors (right) for matrix B.", fig.retina=2}
# target matrix
mat_b <- matrix(c(3, 1, 1, 2), 2)
eigen_transform(matrix = mat_b,
  mat = "B",
  0.6, -1.3,
  -1.2, -0.7,
  1.2, -2,
  -2.5, -2.5
)
```

This time, the eigenvectors have an interesting property. 
They are along the major and minor axes of the ellipse (principal axes),
and are perpendicular to each other. 
An ellipse can be thought of as a circle stretched or shrunk along its principal axes 
as shown in Figure [2](#eigenvectorsb-figure), 
and matrix $\mathbf{B}$ transforms the initial circle $\mathbb{X}$ 
by stretching it along $\mathbf{u}_1$ and $\mathbf{u}_2$, the eigenvectors of $\mathbf{B}$.

How come the eigenvectors of $\mathbf{A}$ did not have this property? 
This is because $\mathbf{B}$ is a symmetric matrix. 
A *symmetric matrix* is a matrix that is equal to its transpose. 
Here is an example of a symmetric matrix:

$$
\begin{bmatrix}
  5 & 0 & 4 & 3 \\
  0 & 7 & 9 & 2 \\
  4 & 9 & 6 & 1 \\
  3 & 2 & 1 & 3 \\
\end{bmatrix}
$$ 

Elements on the main diagonal of a symmetric matrix are arbitrary, 
but for the other elements, each element on row $i$ and column $j$ 
is equal to the element on row $j$ and column $i$ ($a_\text{ij} = a_\text{ji}$). 
A symmetric matrix is always a square matrix ($n \times n$). 
Clearly, $\mathbf{A}$ was not symmetric. 
A symmetric matrix transforms a vector 
by stretching or shrinking the vector along the eigenvectors of this matrix.
In particular, a matrix transforms its eigenvector 
by multiplying its length (or magnitude) by the corresponding eigenvalue. 

Given that the initial vectors in $\mathbb{X}$ all have a length of 1 
and that both $\mathbf{u}_1$ and $\mathbf{u}_2$ are normalized, 
they are members of $\mathbb{X}$. 
Their transformed vectors are: 

$$
\begin{equation*}
  \mathbf{B}\mathbf{u}_1 = {\lambda}_1\mathbf{u}_1,~\\
  \mathbf{B}\mathbf{u}_2 = {\lambda}_2\mathbf{u}_2,
\end{equation*}
$$ 

Therefore, the amount of stretching or shrinking along each eigenvector 
is proportional to the corresponding eigenvalue as shown in Figure [2](#eigenvectorsb-figure).

When you have more stretching in the direction of an eigenvector,
the eigenvalue corresponding to that eigenvector will be greater.
In fact, if the absolute value of an eigenvector is greater than 1,
the circle ð• stretches along it; vice versa.

Let's try another matrix:

$$
C =
\begin{bmatrix}
  3 & 1 \\
  1 & 0.8 \\
\end{bmatrix}
$$


The eigenvectors and corresponding eigenvalues are:

$$
\mathbf{u}_1 = 
\begin{bmatrix}
  0.9327 \\
  0.3606 \\
\end{bmatrix},~\\
\mathbf{u}_2 = 
\begin{bmatrix}
  -0.3606 \\
  0.9327 \\
\end{bmatrix}
$$

$$
\begin{align}
\lambda_1 = 3.3866, \\
\lambda_2 = 0.4134
\end{align}
$$ 

Now if we plot the transformed vectors, we get: 

```{r eigenvectors-matrix-c, class="figure", caption="Eigenvectors (left) and transformed eigenvectors (right) for matrix C", alt="eigenvectorsc. Eigenvectors (left) and transformed eigenvectors (right) for matrix B.", fig.retina=2}
# target matrix
mat_c <- matrix(c(3, 1, 1, 0.8), 2)
eigen_transform(matrix = mat_c,
  mat = "C",
  0.6, -1.3,
  -1.2, -0.7,
  1.2, -0.7,
  -2.5, -2
)
```


Now we have stretched $\mathbf{x}$ along $\mathbf{u}_1$ and shrunk along $\mathbf{u}_2$.

In the [next post]({{< relref "../svd5/index.md" >}}),
we will continue discussing eigenvectors,
but in the context of a *basis* for a vector space.

