---
slug: "srvyr"
title: "Analyzing Survey Data"
date: 2022-01-07T15:09:36-05:00 
publishdate: 2022-01-07
lastmod: 2023-05-21
tags: ["function", "purrr", "srvyr"]
draft: false
autonumbering: true
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
# Options to have images saved in the post folder
# And to disable comment symbols before each line of output
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.retina = 2,
  out.width = "100%",
  fig.dim = c(7, 5),
  results = "hold",
  fig.show = "hold",
  fig.path = "",
  comment = "")

options(tinytex.latexmk.warning = FALSE)

# Knitr hook to use Hugo markdown render image
knitr::knit_hooks$set(plot = function(x, options) {
      base <- knitr::opts_knit$get("base.url")
      if (is.null(base)) base <- ""
      alt <- ifelse(is.null(options$alt), "", options$alt)
      cap <- ifelse(is.null(options$caption), "", options$caption)
      if (alt != "") {
        sprintf('![%s](%s%s "%s")', alt, base, x, cap)
      } else {
        sprintf("![%s](%s%s)", cap, base, x)
        }
  })

# knitr hook to use Hugo highlighting options
knitr::knit_hooks$set(
  source = function(x, options) {
  hlopts <- options$hlopts
    paste0(
      "```r ",
      if (!is.null(hlopts)) {
      paste0("{",
        glue::glue_collapse(
          glue::glue("{names(hlopts)} = {hlopts}"),
          sep = ","
        ), "}"
        )
      },
      "\n", glue::glue_collapse(x, sep = "\n"), "\n```\n"
    )
  }
)
```

```{r include=FALSE}
here::i_am("index.Rmd")
source("../../../static/r/survey.R")
```

```{r include=FALSE}
library(patchwork)
```

This blog is the first one of a series that documents
what I have learned at work while analyzing data from a few national surveys lately.
Instead of using the actual survey data,
I have simulated them to demonstrate the core problem
while keeping the unnecessary details minimal.

## Home ownership

A household well-being survey was administered every couple of years
in two provinces, Ascington and Braeq [^1].
One question on the survey was "Do you currently own or rent your home?"
Figure [1](#housing-figure) shows the results from the most recent two waves of the survey.

[^1]: factitious names generated using 
[this website](https://www.fantasynamegenerators.com/country_names.php).

```{r sim-data, include=FALSE}
set.seed(4)

if (!file.exists(here::here("sim_survey.rds"))) {
  sim_survey <- fabricatr::fabricate(
    province = fabricatr::add_level(N = 2),
    year = fabricatr::add_level(N = 2, nest = FALSE),
    id = fabricatr::link_levels(
      N = 5000,
      by = fabricatr::join_using(province, year),
      housing = sample(c(0, 1), size = 5000, replace = TRUE, prob = c(0.4, 0.6)),
      weight = runif(N, min = 0.5, max = 2.5) # pseudo weight, should add up to 5000
    )
  ) |>
    dplyr::mutate(
      housing = factor(housing,
        levels = c(0, 1),
        labels = c("renter", "owner")
      ),
      province = factor(province,
        levels = c(1, 2),
        labels = c("ascington", "braeq")
      ),
      year = factor(year,
        levels = c(1, 2),
        labels = c("2019", "2021")
      )
    )
    readr::write_rds(sim_survey, here::here("sim_survey.rds"))
} else {
  sim_survey <- readr::read_rds(here::here("sim_survey.rds"))
}

sim_survey_srvy <- sim_survey |>
  srvyr::as_survey(weights = weight)
```


```{r contigency-table}
var_list <- c("housing")
ownership <- cookie_cutter(sim_survey_srvy, var_list, year, province) |>
  purrr::pluck("housing_summary")
```


```{r}
# It's important to not quote additional grouping variables!!
housing_chisq1 <- annotations(
  sim_survey, var1 = "province", var2 = "housing", year)
# housing_chisq2 <- annotations(sim_survey, year, housing, province)
```

```{r gg-template, include=FALSE}
survey_name <- ggplot2::as_labeller(c(
      "2019" = "2019 survey",
      "2021" = "2021 survey")
  )
# prov_name <- ggplot2::as_labeller(c(
#       "ascington" = "Ascington",
#       "braeq" = "Braeq")
#   )
```



```{r viz-housing, class="figure", alt="housing. House ownership between two provinces in 2019 (left) and in 2021 (right).", caption="Home owners."}

p1 <- ownership |>
  dplyr::filter(housing == "owner") |>
  gg_factory(
    x = province,
    wrap = year,
    annotation_df = housing_chisq1)
p1
```

Starting from left panel, results based on the 2019 survey suggest that
there were more home owners in Braeq than in Ascington.
Moving on to the right panel, by 2021, the gap between the two provinces
seemed to have disappeared, if not reversed.
<!-- This change is mainly driven by an increased percentage of home owners -->
This change is mainly driven by a growing number of home owners
in Ascington over the past two years,
and a slightly decreased percentage of home owners in Braeq
during the same time period.
<!-- as indicated by the bottom two panels. -->

<!-- Remaining questions in this type of analysis: -->
<!-- when I split the data by `year` or `province`, -->
<!-- do I need to calculate new weights for each sub-sample? -->
<!-- I suppose I could also group data on `housing` 
and run a chisq on `year + province`. -->

To understand how Figure [1](#housing-figure) is created,
let's first peak at the data underneath the figure.
If you want to follow along, run `chunk 1` to access the demo data.

```{r include=FALSE}
# export data and upload
if(!file.exists(here::here("../../../static/txt/", "housing.txt"))) {
  dput(sim_survey, here::here("../../../static/txt/", "housing.txt"))
}
```


```{r echo=TRUE, eval=FALSE}
# chunk 1 (run this to access the demo data)
sim_survey <- dget("https://raw.githubusercontent.com/chunyunma/chunyunma-dot-me/main/static/txt/housing.txt")
```

`chunk 2` reduces the original data into a contingency table
based on which Figure [1](#housing-figure) was made.


```{r, echo=TRUE, eval=FALSE}
# chunk 2
sim_survey_srvy <- sim_survey |>
  srvyr::as_survey(weights = weight)

ownership <- sim_survey_srvy |>
  dplyr::group_by(year, province, housing) |>
  dplyr::summarize(perc = srvyr::survey_mean(),
    freq = dplyr::n()) |>
  dplyr::select(-perc_se)

owndership
```

```{r}
ownership
```

<!-- Re-arranging it in a more classic view of contingency table in Table x. -->
Now we are ready to plot.

```{r housing-step1, class="figure", alt="step1. Basic ggplot and polished ggplot.", caption="Home owners."}
step1 <- ggplot2::ggplot(
  data = ownership,
  ggplot2::aes(x = province, y = perc, fill = housing)) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ year)
step1 + p1
```


```{r housing-step1-demo, echo=TRUE, eval=FALSE}
step1 <- ggplot2::ggplot(
  data = ownership,
  ggplot2::aes(x = province, y = perc, fill = housing)) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ year)
```

The data underlying both panels in Figure [2](#step1-figure) are identical.
However, they look different.
Figure [3](#steps-figure) illustrates how to construct the final product from Figure 2 (left)
step by step.

```{r}
step1 <- step1 + ggplot2::guides(fill = "none")
```


```{r eval=FALSE, echo=TRUE}
step2 <- ownership |>
  dplyr::filter(housing == "owner") |>
  ggplot2::ggplot(
    ggplot2::aes(x = province, y = perc)) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ year)

step3 <- step2 +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0, 0)
  )

step4 <- step3 +
  ggplot2::scale_x_discrete(
    labels = c(
      "ascington" = "Ascington",
      "braeq" = "Braeq")
  )

step5 <- step4 +
  ggplot2::labs(x = "", y = "Percentage")


survey_name <- ggplot2::as_labeller(c(
      "2019" = "2019 survey",
      "2021" = "2021 survey")
  )

step6 <- step5 +
  ggplot2::facet_wrap(~ year, labeller = survey_name)

step7 <- step6 +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::percent(perc, accuracy = 0.1)),
    position = ggplot2::position_dodge(width = 0.9), vjust = -0.25)

annotation_df <- sim_survey |>
dplyr::group_by(year) |>
tidyr::nest() |>
dplyr::mutate(model = purrr::map(
    data,
    function(data = data) {
      srvyr::as_survey(.data = data, weights = weight) |>
      srvyr::svychisq(formula = ~ province + housing)
    }),
  tidy = purrr::map(model, broom::tidy)
  ) |>
tidyr::unnest(tidy)

p_star <- function(p) {
  symbol <- dplyr::case_when(
    p <= 0.05 ~ "*",
    TRUE ~ "NS")
  return(symbol)
}

step8 <- step7 +
  ggsignif::geom_signif(
    data = annotation_df,
    xmin = "ascington",
    xmax = "braeq",
    y_position = 0.9,
    vjust = -0.2,
    textsize = 3,
    ggplot2::aes(annotations = p_star(p.value)),
    manual = TRUE
  )
```

```{r housing-viz-steps, class="figure", alt="steps. Breakdown of steps on how to construct the polished plot.", caption="Step-by-step demonstration of how to construct a plot.", fig.dim=c(7,8)}
step2 <- ownership |>
  dplyr::filter(housing == "owner") |>
  ggplot2::ggplot(
    ggplot2::aes(x = province, y = perc)) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ year)

step3 <- step2 +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0, 0)
  )

step4 <- step3 +
  ggplot2::scale_x_discrete(
    labels = c(
      "ascington" = "Ascington",
      "braeq" = "Braeq")
  )

step5 <- step4 +
  ggplot2::labs(x = "", y = "Percentage")

step6 <- step5 +
  ggplot2::facet_wrap(~ year, labeller = survey_name)

step7 <- step6 +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::percent(perc, accuracy = 0.1)),
    position = ggplot2::position_dodge(width = 0.9), vjust = -0.25)

step1 + step2 + step3 + step4 + step5 + step6 + step7 + p1 +
  patchwork::plot_layout(ncol = 2) +
  patchwork::plot_annotation(tag_levels = "1", tag_prefix = "Step ",
    tag_suffix = ":") & # be careful of the different connector here
  ggplot2::theme(plot.tag = ggplot2::element_text(size = 8))
```

To recap, 

```{r eval=FALSE, echo=TRUE}
# contingency-table
ownership <- sim_survey_srvy |>
  dplyr::group_by(year, province, housing) |>
  dplyr::summarize(perc = srvyr::survey_mean(),
    freq = dplyr::n()) |>
  dplyr::select(-perc_se)

# chisq for annotation
annotation_df <- sim_survey |>
dplyr::group_by(year) |>
tidyr::nest() |>
dplyr::mutate(model = purrr::map(
    data,
    function(data = data) {
      srvyr::as_survey(.data = data, weights = weight) |>
      srvyr::svychisq(formula = ~ province + housing)
    }),
  tidy = purrr::map(model, broom::tidy)
  ) |>
tidyr::unnest(tidy)

# plot
ownership |>
  dplyr::filter(housing == "owner") |>
  ggplot2::ggplot(ggplot2::aes(x = province, y = perc)) +
    ggplot2::geom_col() +
    ggplot2::facet_wrap(~ year, labeller = survey_name) +
    ggplot2::scale_x_discrete(
      labels = c(
        "ascington" = "Ascington",
        "braeq" = "Braeq")
      ) +
    ggplot2::scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      limits = c(0, 1),
      expand = c(0, 0)
      ) +
    ggplot2::labs(x = "", y = "Percentage") +
    ggplot2::geom_text(
      ggplot2::aes(label = scales::percent(perc, accuracy = 0.1)),
      position = ggplot2::position_dodge(width = 0.9), vjust = -0.25) +
    ggsignif::geom_signif(
      data = annotation_df,
      xmin = "ascington",
      xmax = "braeq",
      y_position = 0.9,
      vjust = -0.2,
      textsize = 3,
      ggplot2::aes(annotations = p_star(p.value)),
      manual = TRUE
    )
```

## Scale up

A typical household survey does not ask just one question.
More likely several dozens of questions.
Consider the following examples:

- Are you currently employed?
- Do you have any financial depedent?
- Do you have any debt?

All of these questions share the same binary structure
as the home ownership question we have seen before.
Presumably, we could re-use the code we have written for the `housing` variable
and create similar figures for new variables.
Or we could write some function!

<!-- You should consider writing a function -->
<!-- whenever you’ve copied and pasted a block of code more than twice -->



